

Sanity check for ComBat in BladderBatch tutorial.
========================================================

```{r, results='hide', echo=FALSE}
starttime = Sys.time()
```
`r as.character(starttime)`

## Overview
This small report aim to show that the application of ComBat as described in the user guide accompanying the R package sva presented in [Leek et al.](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669) will lead to false results even for random data.

Two sanity checks will be performed
- Swapping real data with random numbers
- Permute the group labels whitinh batches
Except for the above alterations, the batch adjustment and analysis are performed as described in the BladderBatch tutorial with additional plots to assess the results.


## Random Data

```{r, results='hide',  message=FALSE, error=FALSE}
library(sva)
library(bladderbatch)
data(bladderdata)
library(pamr)
library(limma)

pheno = pData(bladderEset)
edata = exprs(bladderEset)
```


Substitute the data with random numbers from a normal distribution.
```{r}
set.seed(100)
edata[,] = rnorm(length(edata), mean=0, sd=1)
```

The phenotypes ("cancer" labels) are not evenly spread across batches. Some batches consists of only one type.
```{r}
print(table(pheno[,3:4]))
```


```{r, tidy=FALSE}

mod0 = model.matrix(~1,data=pheno)

batch = pheno$batch

mod = model.matrix(~as.factor(cancer), data=pheno)

combat_edata = ComBat(dat=edata, batch=batch, 
                      mod=mod, numCovs=NULL, 
                      par.prior=TRUE, prior.plots=FALSE)

pValuesComBat = f.pvalue(combat_edata,mod,mod0)
qValuesComBat = p.adjust(pValuesComBat,method="BH")

```

```{r, dev='svg', fig.width=8, fig.height=8}
hist(pValuesComBat,  main="P-values, Random numbers", breaks=100, xlab="p-value")
```



## Permuted Group Labels


```{r}
source("../../commonscripts/helper_functions.r")
nshuffleddatasets = 10
datamatrices_permuted = list()
labelschanged = vector()
permuted_pvalcounts=list()
for(i in 1:nshuffleddatasets)
{
  permutedlabels = shufflesamplesinbatch( rownames(pheno), pheno$batch, pheno$cancer )
  labelschanged = c(labelschanged,  sum(pheno$cancer != pheno[permutedlabels,"cancer"]  ))
  permuteddata = exprs(bladderEset)[,permutedlabels]
  adjusteddata = as.matrix(ComBat(dat=permuteddata, 
                                  batch= pheno$batch, 
                                  mod=model.matrix(~as.factor(pheno$cancer)),  
                                  numCovs=NULL, 
                                  par.prior=TRUE, 
                                  prior.plots=FALSE))
  this_p = f.pvalue(adjusteddata,mod,mod0)
  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
}

```


```{r pvaluespermuted, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}


plot((1:20)/100, permuted_pvalcounts[[1]][1:20],
     main=paste("P-values, real data permuted labels" ,sep=""),
      xlab="p-value", ylab="Frequency", type="l", lwd=1, 
      ylim=c(0, max(unlist( c(permuted_pvalcounts)))))
for(i in 2:length(permuted_pvalcounts))
{
  lines((1:20)/100,permuted_pvalcounts[[i]][1:20])
}

```




```{r pvaluesleeksva, dev='svg', fig.width=10, fig.height=5, tidy=FALSE}
par(mfrow=c(1, 2))
hist(pValuesComBat,  main="P-values, Random numbers", breaks=100, xlab="p-value")
plot((1:20)/100, permuted_pvalcounts[[1]][1:20],
     main=paste("P-values, real data permuted labels" ,sep=""),
      xlab="p-value", ylab="Frequency", type="l", lwd=1, 
      ylim=c(0, max(unlist( c(permuted_pvalcounts)))))
for(i in 2:length(permuted_pvalcounts))
{
  lines((1:20)/100,permuted_pvalcounts[[i]][1:20])
}

```




### References


Johnson, WE, Rabinovic, A, and Li, C (2007). Adjusting batch effects in microarray expression data using Empirical Bayes methods. Biostatistics 8(1):118-127.

Storey, J. D. and Tibshirani, R. (2003) Proc Natl Acad Sci U S A, 100, 9440-5.

Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.(2012) The sva package for removing batch effects and other unwanted variation in high-throughput experiments. Bioinformatics. 2012 Mar 15;28(6):882-3.
http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669

Smyth, GK (2005). Limma: linear models for microarray data. In: 'Bioinformatics and Computational Biology Solutions
  using R and Bioconductor'. R. Gentleman, V. Carey, S. Dudoit, R. Irizarry, W. Huber (eds), Springer, New York, pages
  397-420.
  
  R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing,
  Vienna, Austria. URL http://www.R-project.org/

  Yihui Xie (2013). knitr: A general-purpose package for dynamic report generation in R. R package version 1.5.

  Yihui Xie (2013) Dynamic Documents with R and knitr. Chapman and Hall/CRC. ISBN 978-1482203530

  Yihui Xie (2013) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and
  Roger D. Peng, editors, Implementing Reproducible Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595
  
  RStudio Team (2012). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL http://www.rstudio.com/.


```{r sessionInfo, comment=""}
sessionInfo()
```

generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .
