
Sanity check for ComBat in the SVA tutorial.
========================================================

```{r, results='hide', echo=FALSE}
starttime = Sys.time()
```
`r as.character(starttime)`

## Overview
This small report aim to show that the application of ComBat as described in the user guide accompanying the R package sva presented in [Leek et al.](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669) will lead to false results even for random data.

One sanity checks will be performed
- Swapping real data with random numbers

The rest of the tutorial will be performed with a few extra plots to inspect the p-values from the different analyses.

Most code are taken directly from the vignette source 'vignettes/sva/inst/doc/sva.Rnw'

```{r}

options(width=65)

library(sva)
library(bladderbatch)
data(bladderdata)
library(pamr)
library(limma)

pheno = pData(bladderEset)

edata = exprs(bladderEset)

```

Now the real data is replaced with random numbers. A batch effect is also added since sva needed a batch effect or  it will report "No significant surrogate variables". This is a simple and not necessarily the most realistic way of adding a batch effect.
```{r tidy=FALSE}
set.seed(100)
edata[,] = rnorm(length(edata))
for(batch in (unique(pheno$batch)))
{
	for(s in 1:dim(edata)[1])
	{
		thisgenesbatcheffect = rnorm(1)
		a= pheno$batch==batch
		edata[s, a] = edata[s, a] + thisgenesbatcheffect
	}
}

```


```{r dev='svg', fig.width=6, fig.height=6}

mod = model.matrix(~as.factor(cancer), data=pheno)

mod0 = model.matrix(~1,data=pheno)

n.sv = num.sv(edata,mod,method="leek")
n.sv

svobj = sva(edata,mod,mod0,n.sv=n.sv)

pValues = f.pvalue(edata,mod,mod0)
qValues = p.adjust(pValues,method="BH")
hist(pValues,breaks=100)

```

```{r dev='svg', fig.width=6, fig.height=6}

modSv = cbind(mod,svobj$sv)
mod0Sv = cbind(mod0,svobj$sv)

pValuesSv = f.pvalue(edata,modSv,mod0Sv)
qValuesSv = p.adjust(pValuesSv,method="BH")
hist(pValuesSv,breaks=100)


```

```{r dev='svg', fig.width=6, fig.height=6}

fit = lmFit(edata,modSv)

contrast.matrix <- cbind("C1"=c(-1,1,0,rep(0,svobj$n.sv)),"C2"=c(0,-1,1,rep(0,svobj$n.sv)),"C3"=c(-1,0,1,rep(0,svobj$n.sv)))
fitContrasts = contrasts.fit(fit,contrast.matrix)

eb = eBayes(fitContrasts)
topTableF(eb, adjust="BH")
hist(eb$p.value,breaks=100)

```


Now comes the ComBat way of adjusting for batch.
```{r dev='svg', fig.width=6, fig.height=6}

batch = pheno$batch

mod = model.matrix(~as.factor(cancer), data=pheno)

combat_edata = ComBat(dat=edata, batch=batch, mod=mod, numCovs=NULL, par.prior=TRUE, prior.plots=FALSE)

pValuesComBat = f.pvalue(combat_edata,mod,mod0)
qValuesComBat = p.adjust(pValuesComBat,method="BH")
hist(pValuesComBat,breaks=100)

```

```{r dev='svg', fig.width=6, fig.height=6}

modBatch = model.matrix(~as.factor(cancer) + as.factor(batch),data=pheno)
mod0Batch = model.matrix(~as.factor(batch),data=pheno)
pValuesBatch = f.pvalue(edata,modBatch,mod0Batch)
qValuesBatch = p.adjust(pValuesBatch,method="BH")
hist(pValuesBatch,breaks=100)
```


Prediction
```{r}

n.sv = num.sv(edata,mod,vfilter=2000,method="leek")
svobj = sva(edata,mod,mod0,n.sv=n.sv,vfilter=2000)

set.seed(12354)
trainIndicator = sample(1:57,size=30,replace=F)
testIndicator = (1:57)[-trainIndicator]

trainData = edata[,trainIndicator]
testData = edata[,testIndicator]

trainPheno = pheno[trainIndicator,]
testPheno = pheno[testIndicator,]

mydata = list(x=trainData,y=trainPheno$cancer)
mytrain = pamr.train(mydata)
table(pamr.predict(mytrain,testData,threshold=2),testPheno$cancer)
```
```{r}

trainMod = model.matrix(~cancer,data=trainPheno)
trainMod0 = model.matrix(~1,data=trainPheno)
trainSv = sva(trainData,trainMod,trainMod0)

fsvaobj = fsva(trainData,trainMod,trainSv,testData)
mydataSv = list(x=fsvaobj$db,y=trainPheno$cancer)
mytrainSv = pamr.train(mydataSv)
table(pamr.predict(mytrainSv,fsvaobj$new,threshold=1),testPheno$cancer)

```




### References


Johnson, WE, Rabinovic, A, and Li, C (2007). Adjusting batch effects in microarray expression data using Empirical Bayes methods. Biostatistics 8(1):118-127.

Storey, J. D. and Tibshirani, R. (2003) Proc Natl Acad Sci U S A, 100, 9440-5.

Leek JT, Johnson WE, Parker HS, Jaffe AE, Storey JD.(2012) The sva package for removing batch effects and other unwanted variation in high-throughput experiments. Bioinformatics. 2012 Mar 15;28(6):882-3.
http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669

Smyth, GK (2005). Limma: linear models for microarray data. In: 'Bioinformatics and Computational Biology Solutions
  using R and Bioconductor'. R. Gentleman, V. Carey, S. Dudoit, R. Irizarry, W. Huber (eds), Springer, New York, pages
  397-420.
  
  R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing,
  Vienna, Austria. URL http://www.R-project.org/

  Yihui Xie (2013). knitr: A general-purpose package for dynamic report generation in R. R package version 1.5.

  Yihui Xie (2013) Dynamic Documents with R and knitr. Chapman and Hall/CRC. ISBN 978-1482203530

  Yihui Xie (2013) knitr: A Comprehensive Tool for Reproducible Research in R. In Victoria Stodden, Friedrich Leisch and
  Roger D. Peng, editors, Implementing Reproducible Computational Research. Chapman and Hall/CRC. ISBN 978-1466561595
  
  RStudio Team (2012). RStudio: Integrated Development for R. RStudio, Inc., Boston, MA URL http://www.rstudio.com/.


```{r sessionInfo, comment=""}
sessionInfo()
```

generation ended `r as.character(Sys.time())`. Time spent `r  as.integer(round(difftime(Sys.time(),starttime, units="mins")))` minutes .
